---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Thomas Fox
thanks: "Code and data are available at: https://github.com/ThomasWilliamFox/toronto_child_care"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(palmerpenguins)
library(here)
```


# Introduction

You can and should cross-reference sections and sub-sections. 

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

Some of our data is of penguins (@fig-totalspots), from @palmerpenguins.


```{r}
#| label: fig-totalspots
#| fig-cap: Total number of child-care spaces by ward
#| echo: false
#| warning: false

# read cleaned child care data
child_care = read_csv(
  file = here("outputs/data/child_care_data.csv"),
  show_col_types = FALSE
)

# read ward names
ward_names = read_csv(
  file = here("outputs/data/ward_names.csv"),
  show_col_types = FALSE
)

# turn ward_names into list for graph substitution with ward numbers 
ward_name_list <- as.list(ward_names$ward_name)

# substitute ward numbers for names
child_care_graph_1 <-
  child_care |>
  mutate(ward = as.character(ward))

for (x in 1:25) {
  child_care_graph_1$ward[child_care_graph_1$ward == x] = ward_name_list[[x]]
} 

# construct graph 
child_care_graph_1 |>
  ggplot(aes(x = ward, y = totspace)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 7, angle = 45, vjust = 1, hjust=1))+
  labs(x = "Ward", y = "Total child care spaces")
```

Some of our data is of money (@fig-totalchildren), from @palmerpenguins.

```{r}
#| label: fig-totalchildren
#| fig-cap: Population under 15 years of age by ward
#| echo: false
#| warning: false

# read cleaned child care data
census_data = read_csv(
  file = here("outputs/data/census_data.csv"),
  show_col_types = FALSE
)

census_graph_population <-
  census_data[c(2:26),c(1:7)] |>
  select(ward, pop_0_to_4, pop_5_to_9, pop_10_to_14)

# substitute ward numbers for names
for (x in 1:25) {
  census_graph_population$ward[census_graph_population$ward == x] = ward_name_list[[x]]
} 

# construct graph 
census_graph_population |>
  ggplot(aes(x = ward, y = pop_0_to_4 + pop_5_to_9 + pop_10_to_14)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 7, angle = 45, vjust = 1, hjust=1))+
  labs(x = "Ward", y = "Population under 15 years old") 
```

Some of our data is of money (@fig-avgincome), from @palmerpenguins.

```{r}
#| label: fig-avgincome
#| fig-cap: Average household income by ward
#| echo: false
#| warning: false

# Isolate part of data set needed
census_graph_income <-
  census_data[c(2:26),c(1:7)] |>
  select(ward, avg_hh_income, med_hh_income)

# Substitute ward numbers for names
for (x in 1:25) {
  census_graph_income$ward[census_graph_income$ward == x] = ward_name_list[[x]]
} 

# Construct graph 
census_graph_income |>
  ggplot(aes(x = ward, y = avg_hh_income)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 7, angle = 45, vjust = 1, hjust=1)) +
  labs(x = "Ward", y = "Average household income")
```

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)


Talk way more about it. 

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details



## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...




\newpage


# References


